Modelling state-switching functional data with hidden Markov models

### 2020-12-17 Report/Sections/results.tex

We then select a model corresponding to the fine-scale observations of acceleration. Similarly to the coarse model, we rely on lag plots and visual inspection to select $N^*=3$ Subdive States. Although $N^*$ is selected heuristically, we test the validity of this model in Section \ref{subsec:model_validation}.
%
In contrast to the coarse-scale observations, the fine-scale acceleration data exhibit significant sinusoidal behaviour. Thus, we transform each fine-scale observation sequence $y_t^*$ into $\z_t$ using Equation (\ref{eqn:z}) with a window size of $h=100$ (two seconds) and a maximum frequency of $\tilde{\omega}=10$ (5 Hz). 
%We select a window size of two seconds since it appears be the smallest window size that captures at least one full period of each ``wiggle" in the acceleration data. 
We then have that $\z_{t,\tilde t^*} = \{\zone_{t,\tilde t^*},\ztwo_{t,\tilde t^*}\}$, where $\zone_{t,\tilde t^*}$ is a three dimensional vector of component-wise average acceleration and $\ztwo_{t,\tilde t^*}$ is a scalar describing the ``wiggliness" of a particular window. Even after transforming the raw acceleration data, there is still strong auto-correlation within each component of $\zone_{t,\tilde t^*}$ prior to fitting the model (see Section 1.1 of the supplementary material). Therefore, we choose a CarHMM as defined in Section 2.2 as the fine-scale model.

@@

All components of $\Zone_{t,\tilde t^*}$ are assumed to be Normally distributed as in Equation (\ref{eqn:carhmm}) for all $t$ and $\tilde t^*$. Each component is assumed to have its own mean and variance parameters, but all components share the same auto-correlation parameter. 
%Given that $X^*_{t,\tilde t^*} = i^*$, the corresponding unknown parameters are then $\mathbf{\mu}_A^{*(\cdot,i^*)} \in \mathbb{R}^3$, $\mathbf{\sigma}_A^{*(\cdot,i^*)} \in \mathbb{R}_{>0}^3$, and $\phi_A^{*(\cdot,i^*)} \in [0,1]$.
Given that $\tilde X^*_{t,\tilde t^*} = i^*$, $\Ztwo_{t,\tilde t^*}$ is assumed to follow a Gamma distribution parameterized by its mean $\mu_W^{*(\cdot,i^*)}$ and standard deviation $\sigma_W^{*(\cdot,i^*)}$. We do not include $\Ztwo_{t,\tilde t^*-1}$ in the distribution of $\Ztwo_{t,\tilde t^*}$ because the auto-correlation evident from the lag plot is not severe and may be explained by subsequent observations occurring within the same subdive state. 
%
%The final model is a hierarchical hidden Markov model with a traditional HMM on the coarse scale. Each fine-scale sequence $y^*_t$ is transformed to $\z$ using Equation (\ref{eqn:z}) and $\Z_{t,\tilde t^*}$ is modelled with a CarHMM.
%We assume $\Zone_{t,\tilde t^*}$ is Normally distributed with auto-correlation while $\Ztwo_{t,\tilde t^*}$ is Gamma distributed without auto-correlation. We also assume that $\Ztwo_{t,\tilde t^*}$ and all components of $\Zone_{t,\tilde t^*}$ are each independent of one another when conditioned on the subdive state. The emission distribution of each subdive state does not depend upon the dive type. 
In total, the parameters to estimate are
%
\begin{gather*}
    \Gamma, \qquad \Gamma^{*} = \{\Gamma^{*(1)},\Gamma^{*(2)}\} \qquad \text{(probability transition matrices)}, \\
    %
    \theta = \{\mu^{(1)},\sigma^{(1)},\mu^{(2)},\sigma^{(2)}\} \qquad \text{($Y$ emission parameters), and} \\
    %
    \theta^* = \{\theta^{*(\cdot,1)},\theta^{*(\cdot,2)},\theta^{*(\cdot,3)}\}  \qquad \text{($\Z$ emission parameters), where} \\
    %
    \theta^{*(\cdot,i^*)} =  \{\mu_A^{*(\cdot,i^*)},\sigma_A^{*(\cdot,i^*)},\phi_A^{*(\cdot,i^*)},\mu_W^{*(\cdot,i^*)},\sigma_W^{*(\cdot,i^*)}\}.
\end{gather*}
%
Recall that $\theta^{*(\cdot,i^*)}$ is the set of parameters describing the distribution of $\Z_{t,\tilde t^*}$ conditioned on $\tilde X^*_{t,\tilde t^*} = i^*$. 
%
We refer to this final model as the \textbf{CarHHMM-DFT} since it includes a CarHMM, hierarchical HMM, and DFT-based transformation. The likelihood of this model is easily calculated using the forward algorithm and can be maximized with respect to the parameters above (see appendix for details). Figure \ref{fig:CarHHMM-DFT} shows the dependence structure of the full CarHHMM-DFT.

@@

The coarse-scale parameter estimates suggest that the killer whale has at least two distinct dive behaviours (see Table \ref{table:emis_dists_CarHHMM-DFT} and Figure \ref{fig:coarse_emis}). 
Dive type 1 corresponds to shorter, shallower dives which likely reflects resting, travelling, and to a lesser extent searching for prey.
On the other hand, dive type 2 is longer, deeper, and may be associated with behaviours such as hunting and socialization \citep{Tennessen:2019b}. No dive in this data set has a maximum depth deeper than 30 meters, and a study by \citet{Wright:2017} of killer whales in Johnstone Strait found that most prey captures occur at depths deeper than 100 meters. However, the killer whale studied here was tagged north of Johnstone Strait in Queen Charlotte Sound, and significant numbers of fish have been observed to be caught near the surface (Fortune and Trites, unpublished data). It is therefore unclear whether any of the dives in this study are successful foraging dives.

@@

We also plot histograms of dive duration corresponding to each dive type in Figure \ref{fig:empirical_dist}. Each observation of dive duration is weighted by the probability that it corresponds to a particular dive type as decoded by the \textit{forward-backward algorithm} \citep{Zucchini:2016}. This results in two histograms -- one corresponding to dive type 1 and another corresponding to dive type 2. Each histogram is then plotted together with the corresponding emission distribution learned by the CarHHMM-DFT. Analogous histograms corresponding to the fine-scale observations are contained in Section 1 of the supplementary material. Our results mostly show that the CarHHMM-DFT explains the data well, but there are some exceptions. In particular, $\Ztwo_{t,\tilde t^*}$ is right-skewed and $\Zone_{t,\tilde t^*}$ has heavy tails corresponding to subdive state 3, indicating the existence of rare events corresponding to exceptionally sudden changes in acceleration of the killer whale. These outliers are potential subjects for future study and may indicate biologically relevant phenomena such as prey capture \citep{Tennessen:2019a}.

@@

The CarHHMM does not model the ``wiggliness" of the acceleration data, so it regularly fails to pick up obvious behavioural changes corresponding to the periodicity shown in Figure \ref{fig:labeled_dives} (see Figure 6 of the supplementary material). This essentially disqualifies the CarHHMM as a viable model for this data set. The pseudoresiduals corresponding to the CarHHMM are also light-tailed and do not fit a Normal distribution (see Figure 16 of the supplementary material).

>>>>

We then select a model corresponding to the fine-scale observations of acceleration. Similarly to the coarse model, we rely on lag plots and visual inspection to select $N^*=3$ subdive states. Although $N^*$ is selected heuristically, we test the validity of this model in Section \ref{subsec:model_validation}.
%
In contrast to the coarse-scale observations, the fine-scale acceleration data exhibit significant sinusoidal behaviour. Thus, we transform each fine-scale observation sequence $y_t^*$ into $\z_t$ using Equation (\ref{eqn:z}) with a window size of $h=100$ (two seconds) and a maximum frequency of $\tilde{\omega}=10$ (5 Hz). 
We then have that $\z_{t,\tilde t^*} = \{\zone_{t,\tilde t^*},\ztwo_{t,\tilde t^*}\}$, where $\zone_{t,\tilde t^*}$ is a three dimensional vector of component-wise average acceleration and $\ztwo_{t,\tilde t^*}$ is a scalar describing the ``wiggliness" of a particular window. Even after transforming the raw acceleration data, there is still strong auto-correlation within each component of $\zone_{t,\tilde t^*}$ prior to fitting the model (see Section 1.1 of the supplementary material). Therefore, we choose a CarHMM as defined in Section 2.2 as the fine-scale model.

@@

All components of $\Zone_{t,\tilde t^*}$ are assumed to be Normally distributed as in Equation (\ref{eqn:carhmm}) and independent of one another when conditioned on the dive types and subdive states. Each component is assumed to have its own mean and variance parameters, but all components share the same auto-correlation parameter. 
%Given that $X^*_{t,\tilde t^*} = i^*$, the corresponding unknown parameters are then $\mathbf{\mu}_A^{*(\cdot,i^*)} \in \mathbb{R}^3$, $\mathbf{\sigma}_A^{*(\cdot,i^*)} \in \mathbb{R}_{>0}^3$, and $\phi_A^{*(\cdot,i^*)} \in [0,1]$.
Given that $\tilde X^*_{t,\tilde t^*} = i^*$, $\Ztwo_{t,\tilde t^*}$ is assumed to follow a Gamma distribution parameterized by its mean $\mu_W^{*(\cdot,i^*)}$ and standard deviation $\sigma_W^{*(\cdot,i^*)}$. We do not include $\Ztwo_{t,\tilde t^*-1}$ in the distribution of $\Ztwo_{t,\tilde t^*}$ because the auto-correlation evident from the lag plot is not severe and may be explained by subsequent observations occurring within the same subdive state. 
%
%The final model is a hierarchical hidden Markov model with a traditional HMM on the coarse scale. Each fine-scale sequence $y^*_t$ is transformed to $\z$ using Equation (\ref{eqn:z}) and $\Z_{t,\tilde t^*}$ is modelled with a CarHMM.
%We assume $\Zone_{t,\tilde t^*}$ is Normally distributed with auto-correlation while $\Ztwo_{t,\tilde t^*}$ is Gamma distributed without auto-correlation. We also assume that $\Ztwo_{t,\tilde t^*}$ and all components of $\Zone_{t,\tilde t^*}$ are each independent of one another when conditioned on the subdive state. The emission distribution of each subdive state does not depend upon the dive type. 
In total, the parameters to estimate are
%
\begin{gather*}
    \Gamma, \qquad \Gamma^{*} = \{\Gamma^{*(1)},\Gamma^{*(2)}\} \qquad \text{(probability transition matrices)}, \\
    %
    \theta = \{\mu^{(1)},\sigma^{(1)},\mu^{(2)},\sigma^{(2)}\} \qquad \text{($Y$ emission parameters), and} \\
    %
    \theta^* = \{\theta^{*(\cdot,1)},\theta^{*(\cdot,2)},\theta^{*(\cdot,3)}\}  \qquad \text{($\Z$ emission parameters), where} \\
    %
    \theta^{*(\cdot,i^*)} =  \{\mu_A^{*(\cdot,i^*)},\sigma_A^{*(\cdot,i^*)},\phi_A^{*(\cdot,i^*)},\mu_W^{*(\cdot,i^*)},\sigma_W^{*(\cdot,i^*)}\}.
\end{gather*}
%
Recall that $\theta^{*(\cdot,i^*)}$ is the set of parameters describing the distribution of $\Z_{t,\tilde t^*}$ conditioned on $\tilde X^*_{t,\tilde t^*} = i^*$. 
%
We refer to this final model as the \textbf{CarHHMM-DFT} since it includes a CarHMM, hierarchical HMM, and DFT-based transformation. The likelihood of this model is easily calculated using the forward algorithm and can be maximized with respect to the parameters above (see appendix for details). Figure \ref{fig:CarHHMM-DFT} shows the dependence structure of the full CarHHMM-DFT.

@@

The coarse-scale parameter estimates suggest that the killer whale has at least two distinct dive behaviours (see Table \ref{table:emis_dists_CarHHMM-DFT} and Figure \ref{fig:coarse_emis}). 
Dive type 1 corresponds to shorter, shallower dives which likely reflects resting, travelling, and to a lesser extent searching for prey.
Dive type 2 is longer, deeper, and may be associated with behaviours such as hunting  \citep{Tennessen:2019b}, but it is unclear whether any of the dives in this study are successful foraging dives. No dive in this data set has a maximum depth deeper than 30 meters and a study by \citet{Wright:2017} of killer whales in Johnstone Strait found that most prey captures occur at depths deeper than 100 meters. However, the killer whale studied here was tagged north of Johnstone Strait in Queen Charlotte Sound, and significant numbers of fish have been observed to be caught near the surface (Fortune and Trites, unpublished data). Dive type 2 could also be associated with behaviours such as socializing that can take place several meters below the surface \citep{Tennessen:2019b}.

@@

We also plot histograms of dive duration corresponding to each dive type in Figure \ref{fig:empirical_dist}. Each observation of dive duration is weighted by the probability that it corresponds to a particular dive type as decoded by the \textit{forward-backward algorithm} \citep{Zucchini:2016}. This procedure results in two histograms -- one corresponding to dive type 1 and another corresponding to dive type 2. Each histogram is then plotted together with the corresponding emission distribution learned by the CarHHMM-DFT. Analogous histograms corresponding to the fine-scale observations are contained in Section 1 of the supplementary material. Our results mostly show that the CarHHMM-DFT explains the data well, but there are some exceptions. In particular, $\Ztwo_{t,\tilde t^*}$ is right-skewed and $\Zone_{t,\tilde t^*}$ has heavy tails corresponding to subdive state 3, indicating the existence of rare events corresponding to exceptionally sudden changes in acceleration of the killer whale. These outliers are potential subjects for future study and may indicate biologically relevant phenomena such as prey capture \citep{Tennessen:2019a}.

@@

The CarHHMM does not model the ``wiggliness" of the acceleration data, so it regularly fails to pick up obvious behavioural changes corresponding to the periodicity shown in Figure \ref{fig:labeled_dives} (see Figure 6 of the supplementary material). These results essentially disqualify the CarHHMM as a viable model for this data set. The pseudoresiduals of acceleration are also light-tailed relative to a Normal distribution (see Figure 16 of the supplementary material).

### 2020-12-17 Report/Sections/simulation.tex

We generate 500 independent data sets using the CarHHMM-DFT as a generative model. Each data set consists of a sequence of 100 curves which we call a sequence of killer whale dives. Each dive can be one of $N=2$ dive types based on a Markov chain with probability transition matrix
%
$$\Gamma = \begin{pmatrix} 0.79 & 0.21 \\ 0.81 & 0.19 \end{pmatrix}.$$
%
Dive duration is Gamma distributed, and the coarse-scale emission parameters are $\mu^{(1)} = 25.7s$, $\sigma^{(1)} = 9.6s$, $\mu^{(2)} = 104.6s$, and $\sigma^{(2)} = 64.7s$. After generating the dive durations for all 100 dives in a data set, dive $t$ is broken into a sequence of $\tilde T^*_t = \lfloor Y_t/2 \rfloor$ two-second windows, where the last $Y_t - 2 \tilde T^*_t$ seconds of each simulated dive are ignored. Each two-second segment is assigned one of $N^*=2$ behaviours according to a fine-scale Markov chain $\tilde X^*_t \equiv \left\{\tilde X^*_{t,1}, \ldots, \tilde X^*_{t,\tilde T^*_t} \right\}$ with probability transition matrices
%
$$\Gamma^{*(1)} = \begin{pmatrix} 0.68 & 0.32 \\ 0.05 & 0.95 \end{pmatrix} \quad \text{ and } \quad \Gamma^{*(2)} = \begin{pmatrix} 0.86 & 0.14 \\ 0.15 & 0.85 \end{pmatrix}$$ 
%
for dive types 1 and 2, respectively.
Instead of generating the raw observations $Y^*_{t,t^*}$, we directly simulate the fine-scale transformed observations $\Z_{t,\tilde t^*} = \left\{\Zone_{t,\tilde t^*}, \Ztwo_{t,\tilde t^*}\right\}$. Recall that $\Zone_{t,\tilde t^*}$ and $\Ztwo_{t,\tilde t^*}$ are independent when conditioned on all dive types and subdive states. Each observation of acceleration $\Zone_{t,\tilde t^*}$ is Normally-distributed with explicit auto-correlation while each observation of wiggliness $\Ztwo_{t,\tilde t^*}$ is Gamma-distributed without explicit auto-correlation. In particular, we set the distributions of $\Zone_{t,\tilde t^*}$ and $\Ztwo_{t,\tilde t^*}$ with the following parameters:
%
\begin{gather*}
    \mu_A^{*(\cdot,1)} = 0.0 s, \enspace \sigma_A^{*(\cdot,1)} = 0.034s, \enspace \phi_A^{*(\cdot,1)} = 0.98, \\
    %
    \mu_A^{*(\cdot,2)} = 0.0 s, \enspace \sigma_A^{*(\cdot,2)} = 0.079s, \enspace \phi_A^{*(\cdot,2)} = 0.87, \\
    %
    \mu_W^{*(\cdot,1)} = 23.3, \quad \sigma_W^{*(\cdot,1)} = 13.0, \\
    %
    \mu_W^{*(\cdot,2)} = 301.2, \quad \sigma_W^{*(\cdot,2)} = 330.1.
\end{gather*}
%
It is not possible to uniquely reconstruct the raw accelerometer data $Y^*$ from $\Z$ alone, but we describe one possible mapping from $\Z$ to $Y^*$ in the appendix. Figure \ref{fig:sim_data} shows one realization of $\Z$ for five dives of one simulated data set along with the the corresponding reconstructed realization of $Y^*$.

@@

The full CarHHMM-DFT is the best performing model of the four candidates since it is the generating model. Its average dive decoding accuracy and the average subdive decoding accuracy are both greater than 0.9. All parameter estimates of fine-scale mean values and probability transition matrices, $\hat \mu^*$, $\hat \Gamma^*$ and $\hat \Gamma$, respectively, have statistically insignificant biases. In addition, the biases of $\hat \sigma$, $\hat \phi$, and $\hat \mu$ are either as small or smaller than the other three candidate models. The empirical standard errors of all parameter estimates ($\hat \theta$, $\hat \Gamma$, $\hat \theta^*$, $\hat \Gamma^*$) are well-approximated by the inverse of the observed Fisher information matrix, although the estimated standard errors tend to be slightly smaller than the empirical standard errors. This is especially true for parameters associated with the wiggliness $\Ztwo_{t,\tilde t^*}$, where the empirical standard error can be up to twice as large as the estimated standard error. See Tables 5 through 9 of the supplementary material for detailed results.

The HHMM-DFT performs similarly to the CarHHMM-DFT in most respects. Its average dive decoding accuracy is comparable to the CarHHMM-DFT while its average subdive decoding accuracy is worse by approximately 5 percentage points. Its parameter estimates are comparable to the CarHHMM-DFT with the notable exception that it greatly overestimates $\sigma_A^{*(\cdot,1)}$ and $\sigma_A^{*(\cdot,2)}$. In addition, the estimated standard errors of $\hat \mu_A^{*(\cdot,1)}$, $\hat \mu_A^{*(\cdot,2)}$, $\hat \sigma_A^{*(\cdot,1)}$, and $\hat \sigma_A^{*(\cdot,2)}$ are much smaller than the associated empirical standard errors (see Table 7 of the supplementary material). These results suggest that the estimates of standard deviation can be too large and estimates of standard errors can be too small when auto-correlation is ignored. This finding is consistent with the results of the case study, where the HHMM-DFT produced larger estimates of $\sigma_A^*$ and smaller estimates of standard error compared to the CarHHMM-DFT.

@@

Finally, the CarHMM-DFT is nearly identical to the CarHHMM-DFT in terms of average subdive decoding accuracy, fine-scale parameter biases, and both estimated and empirical standard error for the fine-scale parameter estimates. In addition, the time required to fit the CarHMM-DFT is less than half of that of the other models (see Table \ref{table:accuracy}). However, this model cannot estimate dive type as it lacks any hierarchical structure. The CarHMM-DFT nonetheless fits a (misspecified) single Gamma distribution over the dive duration of all dives. The resulting parameter estimates ($\hat \mu$ and $\hat \sigma$) are highly correlated (See Figure 24 of the supplementary material).

>>>>

We generate 500 independent data sets using the CarHHMM-DFT as a generative model. Each data set consists of a sequence of 100 curves which we call a sequence of killer whale dives. Each dive can be one of $N=2$ dive types based on a Markov chain with probability transition matrix
%
$$\Gamma = \begin{pmatrix} 0.79 & 0.21 \\ 0.81 & 0.19 \end{pmatrix}.$$
%
Dive duration is Gamma distributed, and the coarse-scale emission parameters are $\mu^{(1)} = 25.7s$, $\sigma^{(1)} = 9.6s$, $\mu^{(2)} = 104.6s$, and $\sigma^{(2)} = 64.7s$. After generating the dive durations for all 100 dives in a data set, dive $t$ is broken into a sequence of $\tilde T^*_t = \lfloor Y_t/2 \rfloor$ two-second windows, where the last $Y_t - 2 \tilde T^*_t$ seconds of each simulated dive are ignored. Each two-second segment is assigned one of $N^*=2$ behaviours according to a fine-scale Markov chain $\tilde X^*_t \equiv \left\{\tilde X^*_{t,1}, \ldots, \tilde X^*_{t,\tilde T^*_t} \right\}$ with probability transition matrices
%
$$\Gamma^{*(1)} = \begin{pmatrix} 0.68 & 0.32 \\ 0.05 & 0.95 \end{pmatrix} \quad \text{ and } \quad \Gamma^{*(2)} = \begin{pmatrix} 0.86 & 0.14 \\ 0.15 & 0.85 \end{pmatrix}$$ 
%
for dive types 1 and 2, respectively.
Instead of generating the raw observations $Y^*_{t,t^*}$, we directly simulate the fine-scale transformed observations $\Z_{t,\tilde t^*} = \left\{\Zone_{t,\tilde t^*}, \Ztwo_{t,\tilde t^*}\right\}$. Recall that $\Zone_{t,\tilde t^*}$ and $\Ztwo_{t,\tilde t^*}$ are independent when conditioned on all dive types and subdive states. Each observation of acceleration $\Zone_{t,\tilde t^*}$ is Normally-distributed with explicit auto-correlation while each observation of wiggliness $\Ztwo_{t,\tilde t^*}$ is Gamma-distributed without explicit auto-correlation. In particular, we set the distributions of $\Zone_{t,\tilde t^*}$ and $\Ztwo_{t,\tilde t^*}$ with the following parameters:
%
\begin{gather*}
    \mu_A^{*(\cdot,1)} = 0.0 s, \enspace \sigma_A^{*(\cdot,1)} = 0.034s, \enspace \phi_A^{*(\cdot,1)} = 0.98, \\
    %
    \mu_A^{*(\cdot,2)} = 0.0 s, \enspace \sigma_A^{*(\cdot,2)} = 0.079s, \enspace \phi_A^{*(\cdot,2)} = 0.87, \\
    %
    \mu_W^{*(\cdot,1)} = 23.3, \quad \sigma_W^{*(\cdot,1)} = 13.0, \\
    %
    \mu_W^{*(\cdot,2)} = 301.2, \quad \sigma_W^{*(\cdot,2)} = 330.1.
\end{gather*}
%
It is not possible to uniquely reconstruct the raw accelerometer data $Y^*$ from $\Z$ alone, but we describe one possible mapping from $\Z$ to $Y^*$ in the appendix. Figure 21 of the supplementary material shows one realization of $\Z$ for five dives of one simulated data set along with the the corresponding reconstructed realization of $Y^*$.

@@

The full CarHHMM-DFT is the best performing model of the four candidates since it is the generating model. Its average dive decoding accuracy and the average subdive decoding accuracy are both greater than 0.9. All parameter estimates of fine-scale mean values and probability transition matrices, $\hat \mu^*$, $\hat \Gamma^*$ and $\hat \Gamma$, respectively, have statistically insignificant biases. In addition, the biases of $\hat \sigma$, $\hat \phi$, and $\hat \mu$ are either as small or smaller than the other three candidate models. The empirical standard errors of all parameter estimates ($\hat \theta$, $\hat \Gamma$, $\hat \theta^*$, $\hat \Gamma^*$) are well-approximated by the inverse of the observed Fisher information matrix, although the estimated standard errors tend to be slightly smaller than the empirical standard errors. This small bias is especially noticeable for parameters associated with the wiggliness $\Ztwo_{t,\tilde t^*}$, where the empirical standard error can be up to twice as large as the estimated standard error. See Tables 5 through 9 of the supplementary material for detailed results.

The HHMM-DFT performs similarly to the CarHHMM-DFT in most respects. Its average dive decoding accuracy is comparable to the CarHHMM-DFT while its average subdive decoding accuracy is worse by approximately 5 percentage points. Its parameter estimates are comparable to the CarHHMM-DFT with the notable exception that it greatly overestimates $\sigma_A^{*(\cdot,1)}$ and $\sigma_A^{*(\cdot,2)}$. In addition, the estimated standard errors of $\hat \mu_A^{*(\cdot,1)}$, $\hat \mu_A^{*(\cdot,2)}$, $\hat \sigma_A^{*(\cdot,1)}$, and $\hat \sigma_A^{*(\cdot,2)}$ are much smaller than the associated empirical standard errors (see Table 7 of the supplementary material). These results suggest that the estimates of standard deviation can be too large and estimates of standard errors can be too small when auto-correlation is ignored. This finding is consistent with the results of the case study, where the HHMM-DFT produced larger estimates of $\sigma_A^*$ and smaller estimates of standard error compared to the CarHHMM-DFT. When standard errors are underestimated, the associated confidence intervals are too narrow, implying that researchers may be overconfident in their parameter estimates.

@@

Finally, the CarHMM-DFT is nearly identical to the CarHHMM-DFT in terms of average subdive decoding accuracy, fine-scale parameter biases, and both estimated and empirical standard error for the fine-scale parameter estimates. In addition, the time required to fit the CarHMM-DFT is less than half of that of the other models (see Table \ref{table:accuracy}). However, this model cannot estimate dive type as it lacks any hierarchical structure. The CarHMM-DFT nonetheless fits a (misspecified) single Gamma distribution over the dive duration of all dives. The resulting parameter estimates ($\hat \mu$ and $\hat \sigma$) are highly correlated (See Figure 25 of the supplementary material).

### 2020-12-17 Report/main.tex

% Abstract, keywords, and classification codes
\startabstract{%
\keywords{
\KWDtitle{Key words and phrases}
Multi-state modelling\sep hierarchical modelling\sep time series\sep  biologging\sep statistical ecology\sep animal movement.
% MSC 2010 subject classification codes
\KWDtitle{MSC 2010}Primary 62M05 \sep secondary 62P12}
\begin{abstract}
\abstractsection{}
\abstractsection{Abstract}
%Insert your abstract here; it should
%typically be up to ten lines long. Avoid symbols as much as possible. Formulas
%are strongly discouraged, and citations should be avoided. 
%The title and the abstract should be concise and descriptive;
%list the key words in alphabetical order.  The MSC 2010 subject classification codes can be found here:
%http://www.ams.org/mathscinet/msc/pdfs/classifications2010.pdf.

@@

Data sets comprised of sequences of curves sampled at high frequencies in time are increasingly common in practice, but they can exhibit complicated dependence structures that cannot be modelled using common methods of Functional Data Analysis (FDA). We detail a hierarchical approach which treats the curves as observations from a hidden Markov model (HMM). The distribution of each curve is then defined by another fine-scale model which may involve a variety of methods including moving-window transformations, Fourier analysis, and auto-regression. This approach is broadly applicable to sequences of curves exhibiting intricate dependence structures. As a case study, we use this framework to model the fine-scale kinematic movement of a northern resident killer whale off the western coast of Canada. Through simulations we show that our model produces more interpretable state estimation and more accurate parameter estimates compared to existing baselines.

@@

% Email address for corresponding author
%\correspondingauthor[*]{\\\email{}} % only put this in after acceptance
%
\section{INTRODUCTION}
\input Sections/intro.tex
%
\section{MODELS AND PARAMETER ESTIMATION}
\label{sec:models}
\input Sections/models.tex
%
\section{KILLER WHALE CASE STUDY}
\label{sec:data}
\input Sections/results.tex
%
\section{SIMULATION STUDY}
\label{sec:sim_study}
\input Sections/simulation.tex
%
\section{DISCUSSION}
\input Sections/discussion.tex
%
\begin{ack}{ACKNOWLEDGEMENTS}
This research was enabled in part by support provided by WestGrid (www.westgrid.ca) and Compute Canada (www.computecanada.ca).
\end{ack}
%
\newpage
\bibliography{references}
%
\newpage
\begin{appendix}
\input Sections/appendix.tex
\end{appendix}
%
\newpage
\section{FIGURES AND TABLES}
\input Sections/figures.tex
%
%\CJShistory

>>>>

% Abstract, keywords, and classification codes
\startabstract{%
\keywords{
\KWDtitle{Key words and phrases}
Biologging\sep statistical ecology\sep animal movement\sep accelerometer data\sep diving behaviour\sep killer whales \sep multi-state modelling\sep hierarchical modelling\sep time series.
% MSC 2010 subject classification codes
\KWDtitle{MSC 2010}Primary 62M05 \sep secondary 62P12}
\begin{abstract}
\abstractsection{}
\abstractsection{Abstract}
%Insert your abstract here; it should
%typically be up to ten lines long. Avoid symbols as much as possible. Formulas
%are strongly discouraged, and citations should be avoided. 
%The title and the abstract should be concise and descriptive;
%list the key words in alphabetical order.  The MSC 2010 subject classification codes can be found here:
%http://www.ams.org/mathscinet/msc/pdfs/classifications2010.pdf.

@@

Data sets comprised of sequences of curves sampled at high frequencies in time are increasingly common in practice, but they can exhibit complicated dependence structures that cannot be modelled using common methods of Functional Data Analysis (FDA). We detail a hierarchical approach which treats the curves as observations from a hidden Markov model (HMM). The distribution of each curve is then defined by another fine-scale model which may involve a variety of methods including moving-window transformations, Fourier analysis, and auto-regression. This approach is broadly applicable to sequences of curves exhibiting intricate dependence structures. As a case study, we use this framework to model the fine-scale kinematic movement of a northern resident killer whale off the western coast of Canada. Through simulations, we show that our model produces more interpretable state estimation and more accurate parameter estimates compared to existing baselines.

@@

% Email address for corresponding author
%\correspondingauthor[*]{\\\email{}} % only put this in after acceptance
%
\section{INTRODUCTION}
\input Sections/intro.tex
%
\section{MODELS AND PARAMETER ESTIMATION}
\label{sec:models}
\input Sections/models.tex
%
\section{KILLER WHALE CASE STUDY}
\label{sec:data}
\input Sections/results.tex
%
\section{SIMULATION STUDY}
\label{sec:sim_study}
\input Sections/simulation.tex
%
\section{DISCUSSION}
\input Sections/discussion.tex
%
\begin{ack}{ACKNOWLEDGEMENTS}
%This research was enabled in part by support provided by WestGrid (www.westgrid.ca) and Compute Canada (www.computecanada.ca).
\end{ack}
%
\newpage
\bibliography{references}
%
\newpage
\begin{appendix}
\input Sections/appendix.tex
\end{appendix}
%
\newpage
\section{FIGURES AND TABLES}
\input Sections/figures.tex
%
%\CJShistory
